---
title: "Binary Edge Weight Model"
output:
  rmarkdown::github_document
---

This example covers fitting an edge weight model to presence/absence (binary) data with an observation-level location effect.

*Note: Many of the procedures presented here are stochastic, and plots and results may vary between compilations of this document. In particular, MCMC chains and model estimates may sometimes not be optimal.*

# Setup

First of all we'll load in Rstan for model fitting in Stan, dplyr for handling the data, and igraph for network plotting and computing network centrality. We also load in a custom R file: "simulations.R" to generate synthetic data for this example..

```{r, results='hide', message=FALSE}
library(cmdstanr)
library(dplyr)
library(igraph)

source("../scripts/simulations.R")
```

# Simulating data

Now we will simulate data using the `simulate_binary()` function. The rows of the resulting dataframe describe observations at the dyadic level between nodes. In this dataframe, `event` denotes whether or not an undirected social event was observed in this observation period. The exact definition of observation period will depend on the study, but is commonly a sampling period where at least one of the members of the dyad was observed. This can also be a sampling period where both members of the dyad were observed, and the distinction will affect the interpretation of edge weights. See the paper for further discussion on this. `location` denotes the location at which the observation took place, which may be relevant if location is likely to impact the visibility of social events.

```{r}
# set.seed(123)
data <- simulate_binary()
df <- data$df
head(df)
```

# Preparing the data

Computationally it's easier to work with dyad IDs rather than pairs of nodes in the statistical model, so we'll map the pairs of nodes to dyad IDs before we put the data into the model. The same is true for the location factor, so we will also map the locations to location IDs. We can add these columns to the dataframe using the following code:

```{r, message=FALSE}
df <- df %>%
  group_by(node_1, node_2) %>%
  mutate(dyad_id=cur_group_id()) %>%
  select(node_1, node_2, dyad_id, event, edge_weight_true) %>%
  mutate(event_true=event, edge_weight_true=as.numeric(edge_weight_true))

prob_detect <- 0.6^2
df$event_obs <- rbinom(nrow(df), 1, prob_detect) * df$event_true
head(df)
```
Now we have all of the data in the right format for fitting the model, we just need to put it into a list object. The data required by the statistical model is defined in `binary_model.stan`.

```{r}
model_data <- list(
  N=nrow(df), # Number of observations
  M=max(df$dyad_id), # Number of dyads
  dyad_ids=df$dyad_id, # Vector of dyad IDs corresponding to each observation
  event=df$event_obs # Vector of binary values (0/1, presence/absence) corresponding to each observation
)
```

# Fitting the model

The main purpose of this part of the framework is to estimate edge weights of dyads. To fit the model, we first must compile it and load it into memory using the function `stan_model()` and providing the filepath to the model. The working directory will need to be set to the directory of the model for this to work properly.

```{r, warning=FALSE, message=FALSE, results='hide'}
edge_model <- cmdstan_model("../models/binary_recapture.stan")
```

Compiling the model may take a minute or two, but once this is done, the model can be fit using `sampling()`. The argument `cores` sets the number of CPU cores to be used for fitting the model, if your computer has 4 or more cores, it's worth setting this to 4.

```{r}
fit_edge <- edge_model$sample(model_data, parallel_chains=4)
fit_edge$summary()
```

```{r}
df_agg <- df %>%
  group_by(dyad_id) %>%
  summarise(edge_weight_true=mean(edge_weight_true))
```

```{r}
plot(df_agg$edge_weight_true, fit_edge$draws("edge_weight", format="matrix")[sample(1:4000, 1), ])
abline(a=0, b=1)
```

```{r}
fit_edge$summary("prob_detect")
```

# Model checking

The R-hat values provided by Stan indicate how well the chains have converged, with values very close to 1.00 being ideal. Values diverging too far from 1.00 indicate that the posterior samples may be very unreliable, and shouldn't be trusted. The chains can be plotted using Rstan's `traceplot` function to verify this visually:

```{r, message=FALSE}
traceplot(fit_edge)
```

Good R-hat values don't necessarily indicate that the model is performing well, only that the parameter estimates appear to be robust. To check that the model is performing as it should, a predictive check can be used. A predictive check uses the fitted model to make predictions, and compares those predictions to the observed data. The predictions should indicate that the observed data are concordant with the predictions from the model. There are many ways to perform a predictive check, as data can be summarised in many different ways. For the purposes of this example, we'll use a simple density check where the probability distributions of the aggregated event counts are compared against the predictions from the model. Note that this isn't a guarantee that the model predictions are good, only that the predictions have the same event count distribution as the data. Ideally several predictive checks would be used to check the performance of the model.

This check uses predictions generated by the Stan model as the quantity `event_pred`, with one set of predictions for each step in the MCMC chain. The predictive check will randomly sample 10 of these steps, compute the event counts for each dyad, and plot the densities against the density of the observed event counts from the data.

```{r, message=FALSE}
# Extract event predictions from the fitted model
event_pred <- rstan::extract(fit_edge)$event_pred
num_iterations <- dim(event_pred)[1]

# Plot the density of the observed event counts
plot(density(df_agg$event_count), main="", xlab="Dyadic event counts", ylim=c(0, 0.04))

# Plot the densities of the predicted event counts, repeat for 10 samples
df_copy <- df
for (i in 1:20) {
  df_copy$event <- event_pred[sample(1:num_iterations, size=1), ]
  df_agg_copy <- df_copy %>% 
    group_by(node_1, node_2) %>%
    summarise(event_count=sum(event))
  lines(density(df_agg_copy$event_count), col=rgb(0, 0, 1, 0.5))
}
```

This plot shows that the observed data falls well within the predicted densities, and the predictions suggest the model has captured the main features of the data well. Now we can be reasonably confident that the model has fit correctly and describes the data well, so we can start to make inferences from the model.

# Extracting edge weights

We can access the edge weights using the `logit_p` quantity. This will give a distribution of logit-scale edge weights for each dyad, akin to an edge list. We'll apply the logistic function `plogis` to get the edge weights back to their original scale:

```{r}
logit_edge_samples <- rstan::extract(fit_edge)$logit_edge # Logit scale edge weights
edge_samples <- plogis(logit_edge_samples) # (0, 1) scale edge weights
```

We can summarise the distribution over edge lists by calculating the credible intervals, indicating likely values for each edge. We'll use the 95% credible interval in this example, but there's no reason to choose this interval over any other. The distribution over edge lists can be summarised in the following code:

```{r}
dyad_name <- do.call(paste, c(df_agg[c("node_1", "node_2")], sep=" <-> "))
edge_lower <- apply(edge_samples, 2, function(x) quantile(x, probs=0.025))
edge_upper <- apply(edge_samples, 2, function(x) quantile(x, probs=0.975))
edge_median <- apply(edge_samples, 2, function(x) quantile(x, probs=0.5))
edge_list <- cbind(
  "median"=round(edge_median, 3), 
  "2.5%"=round(edge_lower, 3), 
  "97.5%"=round(edge_upper, 3)
)
rownames(edge_list) <- dyad_name
edge_list
```

In social network analysis, a more useful format for network data is usually adjacency matrices, rather than edge lists, so now we'll convert the distribution of edge lists to a distribution of adjacency matrices, and store the result in an 8 x 8 x 4000 tensor, as there are 8 nodes and 4000 samples from the posterior. 

```{r}
adj_tensor <- array(0, c(8, 8, 4000))
logit_adj_tensor <- array(0, c(8, 8, 4000))
for (dyad_id in 1:model_data$num_dyads) {
  dyad_row <- df_agg[df_agg$dyad_id == dyad_id, ]
  adj_tensor[dyad_row$node_1_id, dyad_row$node_2_id, ] <- edge_samples[, dyad_id]
  logit_adj_tensor[dyad_row$node_1_id, dyad_row$node_2_id, ] <- logit_edge_samples[, dyad_id]
}
adj_tensor[, , 1] # Print the first sample of the posterior distribution over adjacency matrices
```

The adjacency matrix above corresponds to a single draw of the posterior adjacency matrices. You'll notice the edges have been transformed back to the [0, 1] range from the logit scale using the logistic function. If there are no additional effects (such as location in our case), the transformed edge weights will be probabilities and the median will be approximately the same as the simple ratio index for each dyad. However, when additional effects are included, the transformed values can no longer be interpreted as probabilities, though they will be useful for visualisation and analysis purposes.

# Visualising uncertainty

The aim of our network visualisation is to plot a network where the certainty in edge weights (edge weights) can be seen. To do this we'll use a semi-transparent line around each edge with a width that corresponds to a standardised uncertainty measures. The uncertainty measure will simply be the normalised difference between the 97.5% and 2.5% credible interval estimate for each edge weight. We can calculate this from the transformed adjacency tensor object, generate two igraph objects for the main network and the uncertainty in edges, and plot them with the same coordinates.

```{r}
# Calculate lower, median, and upper quantiles of edge weights. Lower and upper give credible intervals.
adj_quantiles <- apply(adj_tensor, c(1, 2), function(x) quantile(x, probs=c(0.025, 0.5, 0.975)))
adj_lower <- adj_quantiles[1, , ]
adj_mid <- adj_quantiles[2, , ]
adj_upper <- adj_quantiles[3, , ]

# Calculate width of credible intervals.
adj_range <- adj_upper - adj_lower
adj_range[is.nan(adj_range)] <- 0

# Generate two igraph objects, one form the median and one from the standardised width.
g_mid <- graph_from_adjacency_matrix(adj_mid, mode="undirected", weighted=TRUE)
g_range <- graph_from_adjacency_matrix(adj_range, mode="undirected", weighted=TRUE)

# Plot the median graph first and then the standardised width graph to show uncertainty over edges.
coords <- igraph::layout_nicely(g_mid)
plot(g_mid, edge.width=3 * E(g_mid)$weight, edge.color="black",  layout=coords)
plot(g_mid, edge.width=20 * E(g_range)$weight, edge.color=rgb(0, 0, 0, 0.25), 
     vertex.label=c("Rey", "Leia", "Obi-Wan", "Luke", "C-3PO", "BB-8", "R2-D2", "D-O"), 
     vertex.label.dist=4, vertex.label.color="black", layout=coords, add=TRUE)
```

This plot can be extended in multiple ways, for example by thresholding low edge weights to visualise the network more tidily, or by adding halos around nodes to show uncertainty around node centrality, and so on.

# Next Steps

Now the edge weight model has been fitted, the edge weight posteriors can be used in the various types of network analyses shown in this repository. 

Save the edge weights for further analysis.

```{r}
data=list(df=df, df_agg=df_agg, logit_edge_samples=logit_edge_samples)
saveRDS(data, file="../example_data/binary.RData")
```
