---
title: "Dyadic Regression with Stan"
output: html_notebook
---

Dyadic regression is a regression analysis often used in social network analysis to determine factors that may be associated with edge weight, for example, age or sex. In this notebook we will use edge weight posteriors from a previously-run edge weight model from binary data to conduct a dyadic regression of edge weight against dyad type. In our toy example, dyad type will be one of `lifeform-lifeform`, `lifeform-droid`, and `droid-droid`. For this example we will be using a Gaussian approximation of edge weights to maintain uncertainty over the data.

# Setup and Loading the edge weights

We'll load edge weight posteriors in from a previous run model using the `readRDS()` function. These data were generated by the `ewm_binary.Rmd` example in the Github repository.

```{r}
data <- readRDS("../example_data/binary.RData")
df <- data$df
df_agg <- data$df_agg
logit_p_samples <- data$logit_p_samples
adj_tensor <- data$adj_tensor
```

# Defining the model

Now everything else is in place, it's time to define the model. As in Stan, the model will be defined by its log-likelihood function, but since we're using a custom Metropolis sampler directly in R, the log-likelihood function will need to be written directly in R. Writing likelihood functions is beyond the scope of this tutorial, but a good resource on the topic can be found here: https://www.ime.unicamp.br/~cnaber/optim_1.pdf.

The dyadic regression model we'll be using will predict the standardised edge weight (Z-score) using a Gaussian family model where dyad type is the main effect, and multi-membership terms are included as random effects to account for non-independence between edges due to nodes. Priors are set relatively narrow to improve model fit for the purposes of this example, but in any real analysis they should be determined by domain knowledge and predictive checks.

```{r}
library(rstan)
model_dyadic <- stan_model("../models/dyadic_regression.stan")
```

```{r}
y_mu <- apply(qlogis(logit_p_samples), 2, mean)
y_sigma <- apply(qlogis(logit_p_samples), 2, sd)

dyad_types <- matrix(0, 8, 8)
dyad_types[1:4, 1:4] <- "ll"; dyad_types[5:8, 5:8] <- "dd"; dyad_types[1:4, 5:8] <- "ld"; dyad_types[5:8, 1:4] <- "ld"
df_agg$dyad_types <- factor(t(dyad_types)[lower.tri(dyad_types)])

model_data <- list(
  N = 28,
  y_mu = y_mu,
  y_sigma = y_sigma,
  dyadtypes = as.integer(df_agg$dyad_types)
)

fit_dyadic <- sampling(model_dyadic, data=model_data, cores=4)
```
The function has evaluated to a real number, so everything appears to be working okay so far. Now we can use the function `metropolis` from `sampler.R` to fit the model using the provided target function, an initial set of parameters (again, all zeros), and some additional MCMC options. Once the sampler has run, we will print out the top few rows of the chains.

```{r}
summary(fit_dyadic)
```


```{r}
chain <- metropolis(target, rep(0, 14), iterations=100000, thin=100, refresh=10000)
colnames(chain) <- c("Intercept", "b_ll", "b_ld", "b_dd", "log(sigma)", "log(mm_sigma)", sapply(1:8, function(i) paste0("mm_", i)))
head(chain)
```

The acceptance rate is around 0.23, which is the target acceptance rate for this sampler. Deviances from 0.23 could indicate sampling issues, but the converse is not true, so acceptance rate isn't an ideal diagnostic tool. Instead it's best to inspect the traceplots, which can be done using the following code:

```{r}
par(mfrow=c(4, 4), mar=c(1,1,1,1))
for (i in 1:14) {
  plot(chain[, i], type="l")
}
```

**Note: In our runs of the code, these traces usually look good. But the stochastic nature of MCMC and the experimental sampler mean that sometimes the chains may not behave well. If the chains in this notebook have not converged, this is simply an artefact of this stochasticity.**

At this point it's a good idea to do some diagnostic checks, such as predictive checks or residual plots. These are covered separately in the Github repository, so diagnostic checks aren't shown here, but should always be carried out.

# Interpreting the model

Pretending we've now carried out any diagnostics we think are appropriate, it's finally time to answer the scientific questions that led us to conduct the analysis. We can start off by calculating the 89% credible intervals of the model parameters. This can be done using the following code:

```{r}
summary_matrix <- t(apply(chain, 2, function(x) quantile(x, probs=c(0.5, 0.055, 0.945))))
summary_matrix <- round(summary_matrix, 2)
summary_matrix
```

At this point it becomes clear that the regression we've conducted is not the same as what might be expected from standard frequentist regressions, where categories are interpreted relative to a reference category. In many cases there's no desire for a reference category, and its presence can necessitate various forms of post-hoc analysis. This is necessitated by the mathematics of most frequentist models, but this is not the case for Bayesian models. Instead, we can use *contrasts* to calculate the magnitude of differences between categories of interest. Contrasts are easily calculated as the statistic of interest from the posteriors of the model. Namely, if we're interest in whether lifeform-lifeform edges are stronger than droid-droid edges, we simply compute the difference in posteriors between `b_ll` and `b_dd`. This can be done using the following code:

```{r}
b_diff <- chain[, "b_ll"] - chain[, "b_dd"]
b_diff_summary <- round(quantile(b_diff, probs=c(0.5, 0.055, 0.945)), 2)
b_diff_summary
```

This gives us an estimate that lifeform-lifeform edges are generally around ~1.7 standard deviations stronger than droid-droid edges, with an 89% credible interval of around ~(1.2, 2.2).

# Conclusion

This notebook has given a brief overview of how to fit and interpret a dyadic regression model using the Metropolis-Hastings sampler in BISoN. We also have a guide on how to conduct the same type of analysis in Stan using a Gaussian approximation of edge weights, which should yield similar results and may be less prone to model fitting issues.
