---
title: "Count Data Example with INLA"
output:
  rmarkdown::github_document
---

This example covers fitting a edge weight model to count data (where the count of social events per observation is recorded) with an observation-level location effect, basic model checking and diagnostics, visualising networks with uncertainty, calculating probability distributions over network centrality, and propagating network uncertainty into subsequent analyses.

*Note: Many of the procedures presented here are stochastic, and plots and results may vary between compilations of this document. In particular, MCMC chains and model estimates may sometimes not be optimal, even if they were when it was originally written.*

# Setup

First of all we'll load in Rstan for model fitting in Stan, dplyr for handling the data, and igraph for network plotting and computing network centrality. We also load in two custom R files: "simulations.R" to generate synthetic data for this example; and "sampler.R" to allow fitting models with uncertainty over network features respectively.

```{r, results='hide', message=FALSE}
library(INLA)
library(dplyr)
library(igraph)

source("../scripts/simulations.R")
source("../scripts/sampler.R")
```

# Simulating data

Now we will simulate data using the `simulate_count()` function. The rows of the resulting dataframe describe observations at the dyadic level between nodes. In this dataframe, `event_count` denotes the number of social events observed in each observation period. The exact definition of observation period will depend on the study, but is commonly a sampling period where at least one of the members of the dyad was observed. This can also be a sampling period where both members of the dyad were observed, and the distinction will affect the interpretation of edge weights. See the paper for further discussion on this. `location` denotes the location at which the observation took place, which may be relevant if location is likely to impact the visibility of social events.

```{r}
set.seed(1)
data <- simulate_count()
df <- data$df
head(df)
```

# Preparing the data

Computationally it's easier to work with dyad IDs rather than pairs of nodes in the statistical model, so we'll map the pairs of nodes to dyad IDs before we put the data into the model. The same is true for the location factor, so we will also map the locations to location IDs. We can add these columns to the dataframe using the following code:

```{r, message=FALSE}
df <- df %>%
  group_by(node_1, node_2) %>%
  mutate(dyad_id=cur_group_id()) %>%
  mutate(location_id=as.integer(location), duration=as.integer(duration))
head(df)
```

It will also be useful later to aggregate the dataframe at the dyad level, assign dyad IDs corresponding to each dyad, and calculate total event counts for each dyad. We can do this using:

```{r, message=FALSE}
df_agg <- df %>%
  group_by(node_1, node_2) %>%
  summarise(event_count_total=sum(event_count), dyad_id=cur_group_id(), total_duration=sum(duration)) %>%
  mutate(node_1_id=as.integer(node_1), node_2_id=as.integer(node_2))
head(df_agg)
```

To fit the model using INLA, we need to define it in the form of a GLM. BISoN models don't need an intercept (though they can be included in modelling relative effects) so the intercept is excluded from the model by adding a preceding `0` to the formula. Dyad ID is modelled (as a factor) as a fixed effect, which will become our edge weight estimates, and location is modelled as a random effect (also known as a varying intercept). Because this model is for count (events per unit time) data, the GLM equivalent family is the Poisson family. Because events can happen over various total observation times, and we want to model event rates that account for this (rather than total numbers of events), an *offset* term must be included. In BISoN count models, the offset term should be the natural logarithm of total observation time (duration). The model can be built using the code below:

```{r}
prior.fixed <- list(mean=0, prec=1) # specify the priors 
prior.random <- list(prec=list(prior="normal", param=c(0, 1)))

# note that in INLA it uses precision instead of SD (which is 1/SD)
df$dyad_id <- as.factor(df$dyad_id)

fit_edge <- inla(
  event_count ~ 0 + dyad_id + f(location, model="iid", hyper=prior.random) + offset(log(duration)), # model edge weights with offset term for observation effort
  family="poisson", # poisson for frequency/count data
  data=df,
  control.fixed=prior.fixed,
  control.compute=list(config = TRUE)
)
```

# Posterior predictive checks

To check the model has captured important aspects of the data well, we can run a predictive check of the predictions of the model against the observed event counts using the following code.

```{r, message=FALSE}
# Extract samples (including predictor values) from posterior of INLA model
inla_samples <- inla.posterior.sample(20, fit_edge)

# Plot the density of the observed event counts
plot(density(df_agg$event_count_total), main="", xlab="Dyadic event counts")

# Plot the densities of the predicted event counts, repeat for multiple samples
df_copy <- df
for (i in 1:length(inla_samples)) {
  j <- sample(1:length(inla_samples), size=1)
  df_copy$event_count <- rpois(nrow(df_copy), exp(head(inla_samples[[j]]$latent, nrow(df))))
  df_agg_copy <- df_copy %>% 
    group_by(node_1, node_2) %>%
    summarise(event_count_total=sum(event_count))
  lines(density(df_agg_copy$event_count_total), col=rgb(0, 0, 1, 0.5))
}
```

This plot shows that the observed data falls well within the predicted densities, and the predictions suggest the model has captured the main features of the data well. Now we can be reasonably confident that the model has fit correctly and describes the data well, so we can start to make inferences from the model.

# Extracting edge weights

The main purpose of this part of the framework is to estimate edge weights of dyads. We can access these using the `logit_p` quantity. This will give a distribution of logit-scale edge weights for each dyad, akin to an edge list. We'll apply the logistic function `plogis` to get the edge weights back to their original scale:

```{r}
# Extract posterior samples again, this time with enough samples to construct reliable CIs
num_samples <- 1000
inla_samples <- inla.posterior.sample(num_samples, fit_edge)
log_edge_samples <- matrix(0, length(inla_samples), 28)
for (i in 1:length(inla_samples)) {
  log_edge_samples[i, ] <- tail(inla_samples[[i]]$latent, 28)
}
edge_samples <- exp(log_edge_samples) # (0, 1) scale edge weights
```

We can summarise the distribution over edge lists by calculating the credible intervals, indicating likely values for each edge. We'll use the 95% credible interval in this example, but there's no reason to choose this interval over any other. The distribution over edge lists can be summarised in the following code:

```{r}
dyad_name <- do.call(paste, c(df_agg[c("node_1", "node_2")], sep=" <-> "))
edge_lower <- apply(edge_samples, 2, function(x) quantile(x, probs=0.025))
edge_upper <- apply(edge_samples, 2, function(x) quantile(x, probs=0.975))
edge_median <- apply(edge_samples, 2, function(x) quantile(x, probs=0.5))
edge_list <- cbind(
  "median"=round(edge_median, 3), 
  "2.5%"=round(edge_lower, 3), 
  "97.5%"=round(edge_upper, 3)
)
rownames(edge_list) <- dyad_name
edge_list
```

In social network analysis, a more useful format for network data is usually adjacency matrices, rather than edge lists, so now we'll convert the distribution of edge lists to a distribution of adjacency matrices, and store the result in an 8 x 8 x 4000 tensor, as there are 8 nodes and 4000 samples from the posterior. 

```{r}
adj_tensor <- array(0, c(8, 8, num_samples))
log_adj_tensor <- array(0, c(8, 8, num_samples))
for (dyad_id in 1:nrow(df_agg)) {
  dyad_row <- df_agg[df_agg$dyad_id == dyad_id, ]
  adj_tensor[dyad_row$node_1_id, dyad_row$node_2_id, ] <- edge_samples[, dyad_id]
  log_adj_tensor[dyad_row$node_1_id, dyad_row$node_2_id, ] <- log_edge_samples[, dyad_id]
}
adj_tensor[, , 1] # Print the first sample of the posterior distribution over adjacency matrices
```

The adjacency matrix above corresponds to a single draw of the posterior adjacency matrices. You'll notice the edges have been transformed back to the >0 range from the log-scale scale using the exponential function. If there are no additional effects (such as location in our case), the transformed edge weights will be probabilities and the median will be approximately the same as the simple ratio index for each dyad. However, when additional effects are included, the transformed values can no longer be interpreted directly as rates, though they will be useful for visualisation and analysis purposes.

# Visualising uncertainty

The aim of our network visualisation is to plot a network where the certainty in edge weights (edge weights) can be seen. To do this we'll use a semi-transparent line around each edge with a width that corresponds to a uncertainty measures. The uncertainty measure will simply be the difference between the 97.5% and 2.5% credible interval estimate for each edge weight. We can calculate this from the transformed adjacency tensor object, generate two igraph objects for the main network and the uncertainty in edges, and plot them with the same coordinates.

```{r}
# Calculate lower, median, and upper quantiles of edge weights. Lower and upper give credible intervals.
adj_quantiles <- apply(adj_tensor, c(1, 2), function(x) quantile(x, probs=c(0.025, 0.5, 0.975)))
adj_lower <- adj_quantiles[1, , ]
adj_mid <- adj_quantiles[2, , ]
adj_upper <- adj_quantiles[3, , ]

# Calculate width/range of credible intervals.
adj_range <- ((adj_upper - adj_lower))
adj_range[is.nan(adj_range)] <- 0

# Generate two igraph objects, one form the median and one from the width.
g_mid <- graph_from_adjacency_matrix(adj_mid, mode="undirected", weighted=TRUE)
g_range <- graph_from_adjacency_matrix(adj_range, mode="undirected", weighted=TRUE)

# Plot the median graph first and then the width graph to show uncertainty over edges.
coords <- igraph::layout_nicely(g_mid)
plot(g_mid, edge.width=0.1 * E(g_mid)$weight, edge.color="black",  layout=coords)
plot(g_mid, edge.width=E(g_range)$weight, edge.color=rgb(0, 0, 0, 0.25), 
     vertex.label=c("Rey", "Leia", "Obi-Wan", "Luke", "C-3PO", "BB-8", "R2-D2", "D-O"), 
     vertex.label.dist=4, vertex.label.color="black", layout=coords, add=TRUE)
```

This plot can be extended in multiple ways, for example by thresholding low edge weights to visualise the network more tidily, or by adding halos around nodes to show uncertainty around network centrality, and so on.

# Next Steps

Now the edge weight model has been fitted, the edge weight posteriors can be used in the various types of network analyses shown in this repository. 
