---
title: "Nodal Regression with Metropolis"
output:
  rmarkdown::github_document
---

Nodal regression is a regression-based analysis often used in social network analysis to determine factors that may be associated with node-level network properties, for example if age or sex predicts network centrality. In this notebook we will use edge weight posteriors from a previously-run edge weight model to conduct a nodal regression of eigenvector centrality against node type. In our toy example, node type will be either `lifeform` or `droid`.

BISoN adopts a fully Bayesian philosophy, so not only does it use edge weight models to estimate uncertainty over edge weights, but it also propagates this uncertainty through downstream analyses, such as nodal regression in this case. There are several ways uncertainty over edge weights can be included in regression analyses, but in this example we will use a Metropolis-Hastings sampler to sample over both posterior centrality estimates *and* the parameter space to model the joint distribution of posterior centralities.

# Setup and Loading the edge weights

We'll load edge weight posteriors in from a previous run model using the `readRDS()` function. These data were generated by the `ewm_binary.Rmd` example in the Github repository.

```{r}
library(rstan)
library(igraph)
library(tidyverse)

source("../scripts/sampler.R")

data <- readRDS("../example_data/binary.RData")
df <- data$df
df_agg <- data$df_agg
logit_edge_samples <- data$logit_edge_samples
```

# Computing centralities

Before we can use centralities in the regression, we must calculate them over the posterior edge weights. We'll use eigenvector centrality as the centrality measure in this example, but any centrality metric can be used. Centrality posteriors can be calculated using the following code:

```{r}
# Build adjacency tensor (4000 x 8 x 8 array of edge weights)
edge_samples <- plogis(logit_edge_samples)
adj_tensor <- array(0, c(4000, 8, 8))
for (dyad_id in 1:nrow(df_agg)) {
  dyad_row <- df_agg[df_agg$dyad_id == dyad_id, ]
  adj_tensor[, dyad_row$node_1_id, dyad_row$node_2_id] <- edge_samples[, dyad_id]
}

# Calculate centrality and store posterior samples in a matrix
centrality_samples <- matrix(0, 4000, 8)
centrality_samples_std <- matrix(0, 4000, 8)
for (i in 1:4000) {
  g <- graph_from_adjacency_matrix(adj_tensor[i, , ], mode="undirected", weighted=TRUE)
  centrality_samples[i, ] <- eigen_centrality(g)$vector
  centrality_samples_std[i, ] <- (centrality_samples[i, ] - mean(centrality_samples[i, ]))/sd(centrality_samples[i, ])
}
head(centrality_samples) # Unstandardised eigenvector centrality
head(centrality_samples_std) # Standardised eigenvector centrality
```

# Preparing the data

Now that the edge weights are loaded, we need to prepare the data for fitting the regression model. This involves 1) preparing a response matrix of centralities describing the distribution of centralities for edge node, and 2) building a predictor matrix indicating the node type of each node ("lifeform" or "droid").

```{r}
predictor_matrix <- matrix(0, 8, 1)
colnames(predictor_matrix) <- c("node_type")
predictor_matrix[, 1] <- c(1, 1, 1, 1, 2, 2, 2, 2)
predictor_matrix
```

# Defining the model

Now everything else is in place, it's time to define the model. As in Stan, the model will be defined by its log-likelihood function, but since we're using a custom Metropolis sampler directly in R, the log-likelihood function will need to be written directly in R. Writing likelihood functions is beyond the scope of this tutorial, but a good resource on the topic can be found here: https://www.ime.unicamp.br/~cnaber/optim_1.pdf.

The dyadic regression model we'll be using will predict the standardised centrality (Z-score) using a Gaussian family model where dyad type is the main effect, and multi-membership terms are included as random effects to account for non-independence between edges due to nodes. Priors are set relatively narrow to improve model fit for the purposes of this example, but in any real analysis they should be determined by domain knowledge and predictive checks.

```{r}
loglik_nodalreg <- function(params, Y, X, index) {
  ### Define parameters ###
  beta_nodetype <- params[1:2]
  sigma <- exp(params[3]) # Exponential keeps underlying value unconstrained, which is much easier for the sampler.
  
  ### Sample data according to index ###
  y <- Y[index %% dim(Y)[1] + 1, ]
  
  ### Define model ###
  
  target <- 0
  
  ## Linear predictor ##
  # y ~ normal(b_intercept + b_dyadtype + mm[i] + mm[j], sigma)
  target <- target + sum(dnorm(y, mean=beta_nodetype[X[, 1]], sd=sigma, log=TRUE)) # Main model
  # b_dyadtype ~ normal(0, 0.5)
  target <- target + sum(dnorm(beta_nodetype, mean=0, sd=1, log=TRUE))
  # sigma ~ exponential(1)
  target <- target + dexp(sigma, rate=1, log=TRUE)
  
  return(target)
}

# Create the `target` function that evaluates the log-likelihood on the dataset.
target <- function(params, index) loglik_nodalreg(params, centrality_samples_std, predictor_matrix, index)
```

We now have a function `target(params, index)` that gives the log-likelihood of a set of parameters `params` given the data, for a particular sample of the posterior edge weights, `index`. Let's make sure it works on an initial set of parameters all set to zero, for an arbitrary index of the data. Note that the standard deviation parameters are transformed by an exponential in the log-likelihood function, so the sampler will treat them on the log-scale, meaning log(sigma) = 0 is equivalent to sigma = 1.

```{r}
target(rep(0, 3), 1)
```

The function has evaluated to a real number, so everything appears to be working okay so far. Now we can use the function `metropolis` from `sampler.R` to fit the model using the provided target function, an initial set of parameters (again, all zeros), and some additional MCMC options. Once the sampler has run, we will print out the top few rows of the chains.

```{r}
chain <- metropolis(target, rep(0, 3), iterations=200000, thin=100, refresh=10000)
chain[, 3] <- exp(chain[, 3])
colnames(chain) <- c("b_lifeform", "b_droid", "sigma")
head(chain)
```

The acceptance rate is around 0.23, which is the target acceptance rate for this sampler. Deviances from 0.23 could indicate sampling issues, but the converse is not true, so acceptance rate isn't an ideal diagnostic tool. Instead it's best to inspect the traceplots, which can be done using the following code:

```{r}
par(mfrow=c(1, 3), mar=c(1,1,1,1))
for (i in 1:3) {
  plot(chain[, i], type="l")
}
```
**Note: In our runs of the code, these traces usually look good. But the stochastic nature of MCMC and the experimental sampler mean that sometimes the chains may not behave well. If the chains in this notebook have not converged, this is likely an artefact of this stochasticity.**

At this point it's a good idea to do some diagnostic checks, such as predictive checks or residual plots. These are covered separately in the Github repository, so in-depth diagnostic checks aren't shown here, but should always be carried out.

# Posterior predictive checks

We will run a brief diagnostic check by comparing the density of expected edge weights (draws of which are shown in black) against the density of predicted edge weights from the regression model (draws of which are shown in blue).

```{r}
plot(density(centrality_samples_std[1, ]), col=rgb(0, 0, 0, 0.25), ylim=c(0, 0.8), main="Posterior predictive density", xlab="Logit edge weight")
for (i in 1:100) {
  j <- sample(1:2000, 1)
  lines(density(centrality_samples_std[j, ]), col=rgb(0, 0, 0, 0.25))
  lines(density(
    rnorm(8, mean=chain[j, 1:2][predictor_matrix[, 1]], sd=chain[j, 3])
  ),
  col=rgb(0, 0, 1, 0.25))
}
```

Almost all of the expected (black) densities fall within the range expected from the predicted (blue) densities. The distribution of edge weights appears to have been captured well by the regression model. There are many other types of diagnostic check that could be carried out, but we won't go into detail here. See the github repository page for more information.

# Interpreting the model

Assuming we've now carried out any diagnostics we think are appropriate, it's finally time to answer the scientific questions that led us to conduct the analysis. We can start off by calculating the 95% credible intervals of the model parameters. This can be done using the following code:

```{r}
summary_matrix <- t(apply(chain, 2, function(x) quantile(x, probs=c(0.5, 0.025, 0.975))))
summary_matrix <- round(summary_matrix, 2)
summary_matrix
```

At this point it becomes clear that the regression we've conducted is not exactly the same as what might be expected from standard frequentist regressions, where categories are interpreted relative to a reference category. Instead, a parameter is estimated for each category, and we can use *contrasts* to calculate the magnitude of differences between categories of interest. Contrasts are easily calculated as the statistic of interest from the posteriors of the model. Namely, if we're interest in whether lifeform-lifeform edges are stronger than lifeform-droid edges, we simply compute the difference in posteriors between `b_ll` and `b_ld`. This can be done using the following code:

```{r}
b_diff <- chain[, "b_lifeform"] - chain[, "b_droid"]
plot(density(b_diff), main="Posterior difference between dyad types")
abline(v=0, lty=2)
```

```{r}
b_diff_summary <- round(quantile(b_diff, probs=c(0.5, 0.025, 0.975)), 2)
b_diff_summary
```

This gives us an estimate that lifeforms have an eigenvector centrality of around ~1.4 standard deviations higher than droids, with a 95% credible interval of approximately (0.5, 2.3).

# Conclusion

This notebook has given a brief overview of how to fit and interpret a dyadic regression model using the Metropolis-Hastings sampler in BISoN. We also have a guide on how to conduct the same type of analysis in Stan (examples/dyadic_regression.Rmd) using a Gaussian approximation of edge weights, which should yield similar results and may be less prone to model fitting issues, at the cost of being less flexible for some types of model.
