---
title: "Nodal Regression with Stan"
output:
  rmarkdown::github_document
---

Nodal regression is a regression-based analysis often used in social network analysis to determine factors that may be associated with node-level network properties, for example if age or sex predicts network centrality. In this notebook we will use edge weight posteriors from a previously-run edge weight model to conduct a nodal regression of eigenvector centrality against node type. In our toy example, node type will be either `lifeform` or `droid`.

BISoN adopts a fully Bayesian philosophy, so not only does it use edge weight models to estimate uncertainty over edge weights, but it also propagates this uncertainty through downstream analyses, such as nodal regression in this case. There are several ways uncertainty over edge weights can be included in regression analyses, but in this example we will use a normal approximation of network centrality posteriors in a Stan model to propagate uncertainty.

# Setup and Loading the edge weights

We'll load edge weight posteriors in from a previous run model using the `readRDS()` function. These data were generated by the `ewm_binary.Rmd` example in the Github repository.

```{r}
library(rstan)
library(igraph)
library(tidyverse)

data <- readRDS("../example_data/binary.RData")
df <- data$df
df_agg <- data$df_agg
logit_edge_samples <- data$logit_edge_samples
```

# Defining the model

The nodal regression model we'll be using will predict node centrality using a Gaussian family model where node type is the main and only effect. Since node centralities will almost always co-vary, we need to model the joint posterior distributions over network centrality as the response variable in the regression model. This can be achieved by modelling the posterior mean centralities `y_mu` as outcomes of a multivariate normal with a covariance matrix `y_sigma`, calculated from the centrality posteriors. In Stan this looks like:

`centrality_mu ~ multi_normal(predictor, centrality_cov + diag_matrix(rep_vector(square(sigma), N)));`

where `predictor` is the predictor term (like `a + b * x` in simple linear regression). Modelling centralities with a multivariate normal allows the joint uncertainty over centralities to be taken into account by the model. Weakly informative priors are used in this example, but in any real analysis they should be determined by domain knowledge and predictive checks.

```{r}
model_nodal <- stan_model("../models/nodal_regression.stan")
```

# Calculate posterior centralities

To parameterise the multivariate normal approximation, we use the sample mean and covariance matrix, calculated from the posterior centrality samples. But first we need to actually calculate the posterior centralities. We will be using eigenvector centrality in this example, which is bounded between 0 and 1. We plan on using a Gaussian family regression model as they are generally easier to interpret and fit. The response of a Gaussian regression is expected to be unbounded, so we will standardise eigenvector centrality. Centrality posteriors can be calculated and standardised using the followign code:

```{r}
# Build adjacency tensor (4000 x 8 x 8 array of edge weights)
edge_samples <- plogis(logit_edge_samples)
adj_tensor <- array(0, c(4000, 8, 8))
for (dyad_id in 1:nrow(df_agg)) {
  dyad_row <- df_agg[df_agg$dyad_id == dyad_id, ]
  adj_tensor[, dyad_row$node_1_id, dyad_row$node_2_id] <- edge_samples[, dyad_id]
}

# Calculate centrality and store posterior samples in a matrix
centrality_samples <- matrix(0, 4000, 8)
centrality_samples_std <- matrix(0, 4000, 8)
for (i in 1:4000) {
  g <- graph_from_adjacency_matrix(adj_tensor[i, , ], mode="undirected", weighted=TRUE)
  centrality_samples[i, ] <- eigen_centrality(g)$vector
  centrality_samples_std[i, ] <- (centrality_samples[i, ] - mean(centrality_samples[i, ]))/sd(centrality_samples[i, ])
}
head(centrality_samples) # Unstandardised eigenvector centrality
head(centrality_samples_std) # Standardised eigenvector centrality
```

Computing these matrices is essential for model fitting, but it's hard to understand differences in centralities intuitively from these raw values. Instead we can visualise the centralities of each node using the following code, using tidyverse as it simplifies the process:

```{r}
df_wide <- data.frame(centrality_samples_std)
colnames(df_wide) <- 1:8
df_long <- pivot_longer(df_wide, cols=1:8, names_to="node_id", values_to="Centrality")
ggplot(df_long, aes(x=Centrality)) +
  geom_density(fill="#387780", alpha=0.7, size=0.4) +
  facet_grid(rows=vars(as.factor(node_id)), scales="free") +
  labs(x="Eigenvector centrality (standardised)") + 
  theme_void() + 
  theme(strip.text.y=element_text(size=12), axis.text.x = element_text(angle = 0, size=12, debug = FALSE), axis.title.x=element_text(size=12), plot.margin = unit(c(0.2,0.2,0.2,0.2), "cm"))
```

This kind of plot can be helpful to understand the data and diagnose issues later on.

# Compute normal approximation

The posterior centralities can now be characterised by the multivariate normal distribution as an approximation to their true posterior distributions. To highlight the importance of sampling from a multivariate normal rather than a univariate normal, we can plot centrality samples from two nodes against each other to see how they co-vary.

```{r}
plot(centrality_samples_std[, 1], centrality_samples_std[, 2])
```

The joint distribution contains important information that we would ideally like to include in the model. The multivariate normal will allow us to do this. Parameterising the multivariate normal approximation is a relatively simple process and can be done by calculating the sample mean and sample covariance matrix of the posterior centrality samples:

```{r}
centrality_mu <- apply(centrality_samples_std, 2, mean)
centrality_cov <- cov(centrality_samples_std)
```

These quantities will be given to the Stan model as data to model joint posteriors of centrality in the regression. We can run a few quick plots to see how well the approximation is working and the covariance for one of the nodes:

```{r}
centrality_samples_sim <- MASS::mvrnorm(1e5, centrality_mu, centrality_cov)

plot(density(centrality_samples_std[, 1]), lwd=2, main="Estimated standardised centrality vs normal approximation", xlab="Logit edge weight")
lines(density(centrality_samples_sim[, 1]), col=rgb(0, 0, 1, 0.5), lwd=2)
```
If we're happy with the approximation, we can now get ready to fit the model.

# Fit the model

The model can be fitted using the following code:

```{r}
model_data <- list(
  num_nodes = 8, # Number of nodes
  centrality_mu = centrality_mu, # Sample means of logit edge weights
  centrality_cov = centrality_cov, # Sample covariance of logit edge weights
  node_type = c(1, 1, 1, 1, 2, 2, 2, 2) # Integer node types corresponding to "lifeform" and "droid"
)

fit_nodal <- sampling(model_nodal, data=model_data, cores=4)
```

# Diagnostics

Now the model is fitted, we can check the traceplots to make sure the chains have converged.

```{r}
traceplot(fit_nodal)
```

The chains are stationary and well-mixed so we should be able to proceed with model diagnostics.

# Posterior predictive checks

We will run a brief diagnostic check by comparing the density of expected centralities (draws of which are shown in black) against the density of predicted centralities from the regression model (draws of which are shown in blue).

```{r}
params <- rstan::extract(fit_nodal)
plot(density(centrality_samples_std[1, ]), main="Posterior predictive density of responses (standardised centrality)", col=rgb(0, 0, 0, 0.25), ylim=c(0, 1))
for (i in 1:100) {
  j <- sample(1:4000, 1)
  lines(density(centrality_samples_std[j, ]), col=rgb(0, 0, 0, 0.25))
  mu <- params$beta_nodetype[j, model_data$node_type]
  sigma <- centrality_cov + diag(rep(params$sigma[j], 8))
  lines(density(MASS::mvrnorm(1, mu, sigma)), col=rgb(0, 0, 1, 0.25))
}
```

Almost all of the expected (black) densities fall within the range expected from the predicted (blue) densities. The distribution of centralities appears to have been captured reasonably by the regression model, but the regression model indicates a lot of variance around predicted values, indicating potential scope for improving the regression model. There are many other types of diagnostic check that could be carried out, but we won't go into detail here. A useful way to think about this kind of plot is to see whether, if one of the black lines had been blue, it would be possible to distinguish it from the rest of the blue lines. This should indicate whether the predicted densities are markedly different to what is expected. See the github repository page for more information.

# Interpreting the model

Assuming we've now carried out any diagnostics we think are appropriate, it's finally time to answer the scientific questions that led us to conduct the analysis. We can start off by calculating the 95% credible intervals of the model parameters. This can be done using the following code:

```{r}
round(summary(fit_nodal)$summary[1:3, c(1, 4, 8)], 2)
```

At this point it becomes clear that the regression we've conducted is not exactly the same as what might be expected from standard frequentist regressions, where categories are interpreted relative to a reference category. Instead, a parameter is estimated for each category, and we can use *contrasts* to calculate the magnitude of differences between categories of interest. Contrasts are easily calculated as the statistic of interest from the posteriors of the model. Namely, if we're interest in whether lifeform nodes are more central than droid nodes, we simply compute the difference in posteriors between `beta_nodetype[1]` and `beta_nodetype[2]`. This can be done using the following code:

```{r}
params <- rstan::extract(fit_nodal)
beta_diff <- params$beta_nodetype[, 1] - params$beta_nodetype[, 2]
plot(density(beta_diff), main="Posterior difference between node types")
abline(v=0, lty=2)
```

```{r}
beta_diff_summary <- round(quantile(beta_diff, probs=c(0.5, 0.025, 0.975)), 2)
beta_diff_summary
```

This gives us an estimate that lifeforms have an eigenvector centrality of around ~1.4 standard deviations higher than droids, with a 95% credible interval of approximately (0.3, 2.4).

# Conclusion

This notebook has given a brief overview of how to fit and interpret a nodal regression model using a normal approximation of eigenvector centrality in Stan.
