---
title: "Count Data Example"
output:
  rmarkdown::github_document
---

This example covers fitting a edge weight model to count data (where the count of social events per observation is recorded) with an observation-level location effect, basic model checking and diagnostics, visualising networks with uncertainty, calculating probability distributions over network centrality, and propagating network uncertainty into subsequent analyses.

*Note: Many of the procedures presented here are stochastic, and plots and results may vary between compilations of this document. In particular, MCMC chains and model estimates may sometimes not be optimal, even if they were when it was originally written.*

# Setup

First of all we'll load in Rstan for model fitting in Stan, dplyr for handling the data, and igraph for network plotting and computing network centrality. We also load in two custom R files: "simulations.R" to generate synthetic data for this example; and "sampler.R" to allow fitting models with uncertainty over network features respectively.

```{r, results='hide', message=FALSE}
library(rstan)
library(dplyr)
library(igraph)

source("../scripts/simulations.R")
source("../scripts/sampler.R")
```

# Simulating data

Now we will simulate data using the `simulate_count()` function. The rows of the resulting dataframe describe observations at the dyadic level between nodes. In this dataframe, `event_count` denotes the number of social events observed in each observation period. The exact definition of observation period will depend on the study, but is commonly a sampling period where at least one of the members of the dyad was observed. This can also be a sampling period where both members of the dyad were observed, and the distinction will affect the interpretation of edge weights. See the paper for further discussion on this. `location` denotes the location at which the observation took place, which may be relevant if location is likely to impact the visibility of social events.

```{r}
set.seed(1)
data <- simulate_count()
df <- data$df
head(df)
```

# Preparing the data

Computationally it's easier to work with dyad IDs rather than pairs of nodes in the statistical model, so we'll map the pairs of nodes to dyad IDs before we put the data into the model. The same is true for the location factor, so we will also map the locations to location IDs. We can add these columns to the dataframe using the following code:

```{r, message=FALSE}
df <- df %>%
  group_by(node_1, node_2) %>%
  mutate(dyad_id=cur_group_id()) %>%
  mutate(location_id=as.integer(location))
head(df)
```

It will also be useful later to aggregate the dataframe at the dyad level, assign dyad IDs corresponding to each dyad, and calculate total event counts for each dyad. We can do this using:

```{r, message=FALSE}
df_agg <- df %>%
  group_by(node_1, node_2) %>%
  summarise(event_count_total=sum(event_count), dyad_id=cur_group_id()) %>%
  mutate(node_1_id=as.integer(node_1), node_2_id=as.integer(node_2))
head(df_agg)
```

Now we have all of the data in the right format for fitting the model, we just need to put it into a list object. The data required by the statistical model is defined in `count_model.stan`.

```{r}
model_data <- list(
  N=nrow(df), # Number of observations
  M=nrow(df_agg), # Number of dyads
  L=6, # Number of locations
  dyad_ids=df$dyad_id, # Vector of dyad IDs corresponding to each observation
  location_ids=df$location_id, # Vector of location IDs corresponding to each observation
  event_count=df$event_count # Vector of event counts corresponding to each observation
)
```

# Fitting the model

To fit the model, we first must compile it and load it into memory using the function `stan_model()` and providing the filepath to the model. The working directory will need to be set to the directory of the model for this to work properly.

```{r, warning=FALSE, message=FALSE, results='hide'}
model <- stan_model("../models/count_model.stan")
```

Compiling the model may take a minute or two, but once this is done, the model can be fit using `sampling()`. The argument `cores` sets the number of CPU cores to be used for fitting the model, if your computer has 4 or more cores, it's worth setting this to 4.

```{r}
fit <- sampling(model, model_data, cores=4, iter=5000, refresh=500)
```

# Model checking

The R-hat values provided by Stan indicate how well the chains have converged, with values very close to 1.00 being ideal. Values diverging from 1.00 indicate that the posterior samples may be very unreliable, and shouldn't be trusted. The chains can be plotted using Rstan's `traceplot` function to verify this visually:

```{r, message=FALSE}
traceplot(fit)
```

Good R-hat values don't necessarily indicate that the model is performing well, only that the parameter estimates appear to be robust. To check that the model is performing as it should, a predictive check can be used. A predictive check uses the fitted model to make predictions, and compares those predictions to the observed data. The predictions should indicate that the observed data are concordant with the predictions from the model. There are many ways to perform a predictive check, as data can be summarised in many different ways. For the purposes of this example, we'll use a simple density check where the probability distributions of the aggregated event counts are compared against the predictions from the model. Note that this isn't a guarantee that the model predictions are good, only that the predictions have the same event count distribution as the data. Ideally several predictive checks would be used to check the performance of the model.

This check uses predictions generated by the Stan model as the quantity `event_pred`, with one set of predictions for each step in the MCMC chain. The predictive check will randomly sample 10 of these steps, compute the event counts for each dyad, and plot the densities against the density of the observed event counts from the data.

```{r, message=FALSE}
# Extract event predictions from the fitted model
event_pred <- extract(fit)$event_pred
num_iterations <- dim(event_pred)[1]

# Plot the density of the observed event counts
plot(density(df_agg$event_count_total), main="", xlab="Dyadic event counts", ylim=c(0, 0.007))

# Plot the densities of the predicted event counts, repeat for 10 samples
df_copy <- df
for (i in 1:20) {
  df_copy$event_count <- event_pred[sample(1:num_iterations, size=1), ]
  df_agg_copy <- df_copy %>% 
    group_by(node_1, node_2) %>%
    summarise(event_count_total=sum(event_count))
  lines(density(df_agg_copy$event_count_total), col=rgb(0, 0, 1, 0.5))
}
```

This plot shows that the observed data falls well within the predicted densities, and the predictions suggest the model has captured the main features of the data well. Now we can be reasonably confident that the model has fit correctly and describes the data well, so we can start to make inferences from the model.

# Extracting edge weights

The main purpose of this part of the framework is to estimate edge weights of dyads. We can access these using the `logit_p` quantity. This will give a distribution of logit-scale edge weights for each dyad, akin to an edge list. A more useful format for network data is usually adjacency matrices, rather than edge lists, so instead we'll convert the distribution of edge lists to a distribution of adjacency matrices, and store the result in an 8 x 8 x 4000 tensor, as there are 8 nodes and 4000 samples from the posterior. 

```{r}
log_p_samples <- extract(fit)$log_p

adj_tensor <- array(0, c(8, 8, num_iterations))
for (dyad_id in 1:model_data$M) {
  dyad_row <- df_agg[df_agg$dyad_id == dyad_id, ]
  adj_tensor[dyad_row$node_1_id, dyad_row$node_2_id, ] <- log_p_samples[, dyad_id]
}
adj_tensor[, , 1] # Print the first sample of the posterior distribution over adjacency matrices
```

The adjacency matrix above corresponds to a single draw of the posterior adjacency matrices. You'll notice that many of the entries are negative, because the edge weights are on the log scale. These can be transformed back to the original range (>0) using the exponential function. If there are no additional effects (such as location in our case), the transformed edge weights will be probabilities and the median will be approximately the same as the simple ratio index for each dyad. However, when additional effects are included, the transformed values can no longer be interpreted as event rates, though they may be useful for visualisation and analysis purposes. We can use the exponential transform to create an adjacency matrix using the function (`exp()` in base R). This will also map 0 values to , so it will be necessary to set those values back to zero again. This transformation can be achieved using the following code:

```{r}
exp(adj_tensor[, , 1]) * upper.tri(adj_tensor[, , 1])
```

It will be necessary to use this transformation for the visualisations and analyses we have planned, so we'll apply the transformation to the entire tensor:

```{r}
adj_tensor_transformed <- adj_tensor
for (i in 1:dim(adj_tensor)[3]) {
  adj_tensor_transformed[, , i] <- exp(adj_tensor[, , i]) * upper.tri(adj_tensor[, , i])
}
```

# Visualising uncertainty

The aim of our network visualisation is to plot a network where the certainty in edge weights (edge weights) can be seen. To do this we'll use a semi-transparent line around each edge with a width that corresponds to a uncertainty measures. The uncertainty measure will simply be the difference between the 97.5% and 2.5% credible interval estimate for each edge weight. We can calculate this from the transformed adjacency tensor object, generate two igraph objects for the main network and the uncertainty in edges, and plot them with the same coordinates.

```{r}
# Calculate lower, median, and upper quantiles of edge weights. Lower and upper give credible intervals.
adj_quantiles <- apply(adj_tensor_transformed, c(1, 2), function(x) quantile(x, probs=c(0.025, 0.5, 0.975)))
adj_lower <- adj_quantiles[1, , ]
adj_mid <- adj_quantiles[2, , ]
adj_upper <- adj_quantiles[3, , ]

# Calculate width/range of credible intervals.
adj_range <- ((adj_upper - adj_lower))
adj_range[is.nan(adj_range)] <- 0

# Generate two igraph objects, one form the median and one from the width.
g_mid <- graph_from_adjacency_matrix(adj_mid, mode="undirected", weighted=TRUE)
g_range <- graph_from_adjacency_matrix(adj_range, mode="undirected", weighted=TRUE)

# Plot the median graph first and then the width graph to show uncertainty over edges.
coords <- igraph::layout_nicely(g_mid)
plot(g_mid, edge.width=E(g_mid)$weight, edge.color="black",  layout=coords)
plot(g_mid, edge.width=2 * E(g_range)$weight, edge.color=rgb(0, 0, 0, 0.25), 
     vertex.label=c("Rey", "Leia", "Obi-Wan", "Luke", "C-3PO", "BB-8", "R2-D2", "D-O"), 
     vertex.label.dist=4, vertex.label.color="black", layout=coords, add=TRUE)
```

This plot can be extended in multiple ways, for example by thresholding low edge weights to visualise the network more tidily, or by adding halos around nodes to show uncertainty around network centrality, and so on.

# Extracting network centralities

Uncertainty around network metrics such as centrality can be calculated quite simply by drawing adjacency matrices from the posterior distribution over adjacency matrices, generating a network from them, and calculating the network metric of interest. It is important to sample over the adjacency matrices rather than by the edges on their own, as this maintains the joint distribution of edge weights and will generate more reliable and accurate estimates of network centrality.

```{r}
centrality_matrix <- matrix(0, nrow=num_iterations, ncol=8)
for (i in 1:num_iterations) {
  g <- graph_from_adjacency_matrix(adj_tensor[, , i], mode="undirected", weighted=TRUE)
  centrality_matrix[i, ] <- strength(g)
}
colnames(centrality_matrix) <- c("Rey", "Leia", "Obi-Wan", "Luke", "C-3PO", "BB-8", "R2-D2", "D-O")
head(centrality_matrix)
```

Each column in this matrix corresponds to one of the nodes in the network, and each row is its centrality in one sample of the posterior distribution of the adjacency matrices. We can calculate the credible intervals using the `quantile` function as follows:

```{r}
centrality_quantiles <- t(apply(centrality_matrix, 2, function(x) quantile(x, probs=c(0.025, 0.5, 0.975))))
centrality_quantiles
```

# Maintaining uncertainty in regression on centralities

The key challenge to quantifying uncertainty in network analysis is to incorporate uncertainty due to sampling into downstream analyses, commonly regression. This can be achieved by modifying the likelihood function of a regression model to treat the network centralities with uncertainty. We have written a custom MCMC sampler function that samples from the joint distribution of network centralities calculated earlier and treats those samples as the data in the likelihood function. Likelihood functions for the sampler use the `index` variable to keep track of which data points are being compared internally in the sampler, to ensure that candidate steps in the MCMC are not accepted or rejected because they are being compared to different data points, rather than because of the parameter space. 

Custom likelihood functions take a similar form to the `target +=` syntax in Stan, but for more specific resources the following document is a good start: https://www.ime.unicamp.br/~cnaber/optim_1.pdf. We will implement a linear regression to test if lifeforms are more central in the social network than droids. We have included a coefficient for both lifeform and droid, unlike standard frequentist models. This is because using a reference category (such as droid) would imply that there is less uncertainty around the centrality of droids than around lifeforms. It also allows for easy comparison between categories by calculating the difference in posteriors. 

```{r}
loglik <- function(params, Y, X, index) {
  # Define parameters
  intercept <- params[1]
  beta_lifeform <- params[2]
  beta_droid <- params[3]
  sigma <- exp(params[4]) # Exponential keeps underlying value unconstrained, which is much easier for the sampler.
  
  # Sample data according to index
  y <- Y[index %% dim(Y)[1] + 1, ]
  
  # Define model
  target <- 0
  target <- target + sum(dnorm(y, mean=intercept + beta_lifeform * X[, 1] + beta_droid * X[, 2], sd=sigma, log=TRUE)) # Main model
  target <- target + dnorm(intercept, mean=0, sd=2.5, log=TRUE) # Prior on intercept
  target <- target + dnorm(beta_lifeform, mean=0, sd=2.5, log=TRUE) # Prior on lifeform coefficient
  target <- target + dnorm(beta_droid, mean=0, sd=2.5, log=TRUE) # Prior on droid coefficient
  target <- target + dexp(sigma, 1, log=TRUE) # Prior on sigma
  
  return(target)
}
```

Now we will prepare data for fitting the model. The predictor matrix is simply a matrix with 2 columns and 8 rows, corresponding to whether each of the 8 nodes is a lifeform (column 1) or a droid (column 2).

```{r}
predictor_matrix <- matrix(0, nrow=8, ncol=2)
colnames(predictor_matrix) <- c("lifeform", "droid")
predictor_matrix[1:4, 1] <- 1
predictor_matrix[5:8, 2] <- 1
predictor_matrix
```

Since network strength is strictly positive, a Gaussian error is not a reasonable model for the data. The Gaussian family model is much easier to implement as well as interpret than many other models, so we will standardise the centralities by taking z-scores.

```{r}
centrality_matrix_std <- (centrality_matrix - apply(centrality_matrix, 1, mean))/apply(centrality_matrix, 1, sd)
centrality_matrix_std[is.nan(centrality_matrix_std)] <-0
head(centrality_matrix_std)
```

Now we're in a position to fit the model. To do this, we define the target function, which is simply a function that maps candidate parameters and a network centrality index to the log-likelihood of that function for the given sample of the centrality posterior. This means the target function can be written as a function of the data `centrality_matrix_std` and `predictor_matrix`.

```{r}
target <- function(params, index) loglik(params, centrality_matrix_std, predictor_matrix, index)
```

The function `metropolis` from `sampler.R` can now be used to fit the model using the provided target function, an initial set of parameters, and some additional MCMC options.

```{r}
chain <- metropolis(target, c(0, 0, 0, 0), iterations=100000, thin=100, refresh=10000)
colnames(chain) <- c("intercept", "beta_lifeform", "beta_droid", "sigma")
head(chain)
```

# Checking the regression

The resulting chain of MCMC samples forms the posterior distribution of parameter estimates for the regression model. But before we look at these too closely, we should check that the chains have converged:

```{r}
par(mfrow=c(2, 2))
for (i in 1:4) {
  plot(chain[, i], type="l")
}
```

These chains appear to be quite healthy. Ideally we would run multiple additional chains starting at different points to check that they converge and mix properly. For the sake of this example we won't go into that here.

Again, the performance of the sampler doesn't necessarily guarantee the performance of the model, so we'll use predictive checks to test the performance of the model. In this case, the data are not fixed, and there are multiple possible values they can take. Therefore we'll plot the distribution of centrality values on different draws of the adjacency matrices as well as the distribution of predicted centrality values on different draws.

```{r}
plot(density(centrality_matrix_std[1, ]), ylim=c(0, 0.7), main="", xlab="Standardised network strength")
sample_ids <- sample(1:1000, size=200)
for (i in sample_ids) {
  pred <- rnorm(8, mean=chain[i, "intercept"] + chain[i, "beta_lifeform"] * predictor_matrix[, 1] + chain[i, "beta_droid"] * predictor_matrix[, 2], sd=exp(chain[i, "sigma"]))
  lines(density(centrality_matrix_std[i, ]), col=rgb(0, 0, 0, 0.25))
  lines(density(pred), col=rgb(0, 0, 1, 0.25))
}
```

The model appears to fit reasonably well, and the observed data are completely consistent with the predictions of the model, so we can go ahead with the analysis.

# Interpreting the regression

The regression coefficients and parameters can be summarised by calculating their percentile credible interval similar to before:

```{r}
coefficient_quantiles <- t(apply(chain, 2, function(x) quantile(x, probs=c(0.025, 0.5, 0.975))))
coefficient_quantiles
```

A frequentist analysis (and some Bayesian ones too) would have only one category, lifeform or droid, and the other category would be the implicit reference category, absorbed by the intercept. In this type of analysis, the coefficients for the two categories correspond to the average difference between the centrality of nodes in that category compared to the population average (the intercept). Therefore, to look for a difference between the two categories, we can simply calculate the difference in the posterior distributions of those two categories:

```{r}
beta_difference <- chain[, "beta_lifeform"] - chain[, "beta_droid"]
quantile(beta_difference, probs=c(0.025, 0.5, 0.975))
```

The mass of probability is with there being a positive difference of around 1.75 standard deviations between the centralities of lifeforms compared to droids. Many of the benefits of Bayesian analysis only apply when significance testing is avoided. Though it is reasonably common for a result such as the one above not overlapping zero to be interpreted as being "significant", using such a decision rule leaves Bayesian analysis open to the same flaws as frequentist analyses often have. For this reason we caution against using such a rule.

# Conclusion

In this guide we have shown how to apply the edge weight model to count data and how to conduct subsequent analyses, while maintaining uncertainty through the whole process. Though this process is quite hands-on, it provides a huge amount of flexibility for conducting animal social network analyses in a robust and interpretable way.
